---
title: 'Seminar 10: Descriptive statistics'
output: html_notebook
---

## 1. Data (Id is the average of the member's no. in the P&S list)

### Task: Generate your data

This is a sample of size $n=1000$ combined from two samples of size $500$ from the normal distributions $N(\mu_k,\sigma_k^2)$ with $\mu_1 = 43$, $\sigma_1 = 3$ and $\mu_2 = -2$, $\sigma_2 = 43$

Hint: you can use $\texttt{mean = c}(\mu_1,\mu_2)$ and $\texttt{sd = c}(\sigma_1,\sigma_2)$

```{r}
# your code here
n <- 1000
mean1 = 43
sd1 = 3
mean2 = -2
sd2 = 43

sample_data <- rnorm(n, c(mean1, sd1), c(mean2, sd2))

#my.data <- rnorm(...)
#summary(my.data)
```

## 2. Visualise the data

### Task: Draw the histogram of the data, empirical density, and empirical cdf. Comment on whether the data are close to a normal one

```{r}
# draw histogram
hist(sample_data, 
     probability = TRUE, 
     breaks = 30, 
     col = "lightblue",
     main = "Histogram with Empirical and Normal Densities",
     xlab = "Sample values",
)

# superimpose empirical density
lines(density(sample_data), col="red", lwd=2)

# superimpose normal density
x <- seq(min(sample_data), max(sample_data), length = 1000)
lines(x, dnorm(x, mean = 43, sd = 3), col = "darkblue", lwd = 2)
lines(x, dnorm(x, mean = -2, sd = 43), col = "green", lwd = 2)

```

### Plot ecdf

```{r}
# plot ecdf
plot(ecdf(sample_data))

# superimpose cdf for standard normal
lines(x, pnorm(x, mean = 43, sd = 3), col ="red", lwd = 2)
lines(x, pnorm(x, mean = -2, sd = 43), col ="darkblue", lwd = 2)

# calculate maximal difference 
max(abs(ecdf(sample_data)(x)-pnorm(x, mean = 43, sd = 3)))
max(abs(ecdf(sample_data)(x)-pnorm(x, mean = -2, sd = 43)))

```

## 3. Skewness and kurtosis of the data

**Skewness** is defined as $$
  \mathsf{E} \left[\left({\frac {X-\mu }{\sigma }}\right)^{3}\right]={\frac {\mu _{3}}{\sigma ^{3}}}={\frac {\mathsf{E} \left[(X-\mu )^{3}\right]}{\ \ \ (\mathsf{E} \left[(X-\mu )^{2}\right])^{3/2}}}
$$

Positive skewness means the data have longer right tail and then the mean value $\mathsf{E}(X)$ is greater than the median

**Kurtosis** is defined as $$
  \mathsf{E} \left[\left({\frac {X-\mu }{\sigma }}\right)^{4}\right]={\frac {\mu _{4}}{\sigma ^{4}}}={\frac {\mathsf{E} \left[(X-\mu )^{4}\right]}{\ \ \ (\mathsf{E} \left[(X-\mu )^{2}\right])^{2}}}
$$

For a standard normal distribution, the kurtosis is $3$. If we get a larger value, then more data are concentrated near the mean, and the empirical density is more steep

```{r}
# your code here
# skewness
skewness <- (n / ((n - 1) * (n - 2))) * sum(((sample_data - mean(sample_data)) / sd(sample_data))^3)
skewness

# kurtosis
kurtosis <- ((n * (n + 1)) / ((n - 1) * (n - 2) * (n - 3))) * sum(((sample_data - mean(sample_data)) / sd(sample_data))^4) - (3 * (n - 1)^2) / ((n - 2) * (n - 3))
kurtosis

```

## 4. Percentiles

### For q= 0.1,...,0.9 calculate sample q-percentiles manually (i.e. by sorting the data and picking the corresponding values) and by using percentile

```{r}
percentiles <- function(data, method = "manual") {
  
  sorted_data <- sort(data)
  n <- length(sorted_data)
  per <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9)
  
  if (method == "manual") {
    
    manual_percentile <- function(q) {
      index <- q * (n - 1)
      floor_index <- floor(index)
      ceiling_index <- ceiling(index)
      
      if (floor_index == ceiling_index) {
        return(sorted_data[floor_index + 1])
      } else {
        lower_value <- sorted_data[floor_index + 1]
        upper_value <- sorted_data[ceiling_index + 1]
        fraction <- index - floor_index
        return(lower_value + fraction * (upper_value - lower_value))
      }
    }
    
    result <- sapply(per, manual_percentile)
    
  } else if (method == "quantile") {
    result <- quantile(data, per)
    
  } else {
    stop("Invalid method given. Use 'manual' or 'quantile'.")
  }
  
  return(result)
}

manual_percentiles <- percentiles(sample_data, method = "manual")
cat("Manual Percentile Calculation:\n")
print(manual_percentiles)

quantile_percentiles <- percentiles(sample_data, method = "quantile")
cat("Quantile Function Calculation:\n")
print(quantile_percentiles)


```

## 5. Sample mean and sample standard error vs the theoretical ones

### What is the theoretical expected value of the random variable considered? What is the variance? Compare them to the sample values; explain the difference in the variance

```{r}
# your code here
# theoretical expected value and variance
thr_mean =  
thr_var = 

# sample mean and sample variance
sampl_mean = mean(sample_data)
sample_var = var(sample_data)

```

## 6. k-sigma rule

### Calculate the fraction of the data that are within $k\sigma$ of the sample mean for $k=1,2,3$. Do we get the result expected? Why or why not?

```{r}
# your code here

```
