---
title: 'Seminar 10: Descriptive statistics'
output: html_notebook
---

## 1. Data (Id is the average of the member's no. in the P&S list)

### Task: Generate your data

This is a sample of size $n=1000$ combined from two samples of size $500$ from the normal distributions $N(\mu_k,\sigma_k^2)$ with $\mu_1 = Id$, $\sigma_1 = 3$ and $\mu_2 = -2$, $\sigma_2 = Id$

Hint: you can use $\texttt{mean = c}(\mu_1,\mu_2)$ and $\texttt{sd = c}(\sigma_1,\sigma_2)$

```{r}
# your code here
n <- 1000
mu1 <- 50
sigma1 <- 3
mu2 <- -2
sigma2 <- 50

#my.data <- rnorm(n, mean=c(mu1,mu2), sd=c(sigma1,sigma2))

data1 <- rnorm(n/2, mean=mu1, sd=sigma1)
data2 <- rnorm(n/2, mean=mu2, sd=sigma2)
my.data <- c(data1, data2)
summary(my.data)

```

## 2. Visualise the data

### Task: Draw the histogram of the data, empirical density, and empirical cdf. Comment on whether the data are close to a normal one

```{r}
# draw histogram

# superimpose empirical density

# superimpose normal density

```

### Plot ecdf

```{r}
# plot ecdf

# superimpose cdf for standard normal

# calculate maximal difference

n <- 1000
mu_1 <- 0
sigma_1 <- 3
mu_2 <- -2
sigma_2 <- mu_1
set.seed(123)
data_1 <- rnorm(500, mean = mu_1, sd = sigma_1)
data_2 <- rnorm(500, mean = mu_2, sd = sigma_2)

my.data <- c(data_1, data_2)
hist(my.data, breaks = 30, probability = TRUE, 
     main = "Histogram and empirical density", 
     xlab = "Values", col = "lightblue")
lines(density(my.data), col = "red", lwd = 2)
curve(dnorm(x, mean = mean(my.data), sd = sd(my.data)), 
      add = TRUE, col = "green", lwd = 2)
legend("topright", legend = c("Empirical density", "Normal density"), 
       col = c("red", "green"), lwd = 2)

```

## 3. Skewness and kurtosis of the data

**Skewness** is defined as $$
  \mathsf{E} \left[\left({\frac {X-\mu }{\sigma }}\right)^{3}\right]={\frac {\mu _{3}}{\sigma ^{3}}}={\frac {\mathsf{E} \left[(X-\mu )^{3}\right]}{\ \ \ (\mathsf{E} \left[(X-\mu )^{2}\right])^{3/2}}}
$$

Positive skewness means the data have longer right tail and then the mean value $\mathsf{E}(X)$ is greater than the median

**Kurtosis** is defined as $$
  \mathsf{E} \left[\left({\frac {X-\mu }{\sigma }}\right)^{4}\right]={\frac {\mu _{4}}{\sigma ^{4}}}={\frac {\mathsf{E} \left[(X-\mu )^{4}\right]}{\ \ \ (\mathsf{E} \left[(X-\mu )^{2}\right])^{2}}}
$$

For a standard normal distribution, the kurtosis is $3$. If we get a larger value, then more data are concentrated near the mean, and the empirical density is more steep

```{r}
n <- length(my_data)
mean_data <- mean(my_data)
deviation <- my_data - mean_data

second_moment <- sum(deviation^2) / n
third_moment <- sum(deviation^3) / n
fourth_moment <- sum(deviation^4) / n

skewness_manual <- third_moment / (second_moment^(3/2))
kurtosis_manual <- fourth_moment / (second_moment^2)


```

## 4. Percentiles

### For q= 0.1,...,0.9 calculate sample q-percentiles manually (i.e. by sorting the data and picking the corresponding values) and by using percentile


percentiles <- function(data, method = "manual") {
  sorted_data <- sort(data)
  n <- length(sorted_data)
  per <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9)
  
  if (method == "manual") {
    manual_percentile <- function(q) {
      index <- q * (n - 1)
      floor_index <- floor(index)
      ceiling_index <- ceiling(index)
      
      if (floor_index == ceiling_index) {
        return(sorted_data[floor_index + 1])
      } else {
        lower_value <- sorted_data[floor_index + 1]
        upper_value <- sorted_data[ceiling_index + 1]
        fraction <- index - floor_index
        return(lower_value + fraction * (upper_value - lower_value))
      }
    }
    result <- sapply(per, manual_percentile)
    
  } else if (method == "quantile") {
    # Use built-in quantile function
    result <- quantile(data, per)
    
  } else {
    stop("Invalid method given. Use 'manual' or 'quantile'.")
  }
  
  return(result)
}

data <- c(12, 15, 14, 10, 18, 20, 11)

manual_percentiles <- percentiles(data, method = "manual")
cat("Manual Percentile Calculation:\n")
print(manual_percentiles)

quantile_percentiles <- percentiles(data, method = "quantile")
cat("Quantile Function Calculation:\n")
print(quantile_percentiles)

## 5. Sample mean and sample standard error vs the theoretical ones

### What is the theoretical expected value of the random variable considered? What is the variance? Compare them to the sample values; explain the difference in the variance

```{r}
# your code here
# theoretical expected value and variance

# sample mean and sample variance

```

## 6. k-sigma rule

### Calculate the fraction of the data that are within $k\sigma$ of the sample mean for $k=1,2,3$. Do we get the result expected? Why or why not?

```{r}
# your code here

```
